---
output: 
  rmarkdown::html_document :
    keep_md: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = '100%'
)
options(width=100)
quick_build <- T # wether to cache heavy-computation chunks 

library(RAPToR)
library(RColorBrewer)
library(beeswarm)

requireNamespace("limma", quietly = T)
requireNamespace("stats")

transp <- function(col, a=.5){
  colr <- col2rgb(col)
  return(rgb(colr[1,], colr[2,], colr[3,], a*255, maxColorValue = 255))
}
```

### The data

**The datasets in this example are those used for all the *in-text* examples throughout the reference-building vignette.**

We are using the two *C. elegans* time series datasets. 

 1. A time series of larval development in 4 different strains published by @aeschimann2017lin41, called `ds1`. This is the dataset used to build the reference. (Accession : [GSE80157](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE80157))
 1. A high-resolution time series of late larval development published by @hendriks2014extensive, called `ds2`. This is the dataset used for external validation. (Accession : [GSE52861](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE52861))

#### {.tabset}
Code to generate `ds1` and `ds2` :

##### Hide

##### Show
```{r raw2rpkm_func}
raw2rpkm <- function(X, gene.length, id.col = 1, l.col='length'){
  # Compute RPKM from raw counts
  if(!all(rownames(X)%in%gene.length[, id.col])){
    stop("Some genes are missing length info !")
  }
  res <- sapply(colnames(X), function(samp){
    pm <- sum(X[,samp])/1e6
    rpkm <- (X[,samp]/pm)/(gene.length[match(rownames(X), gene.length[, id.col]), l.col]/1000)
  })
  rownames(res) <- rownames(X)
  return(res)
}
```

<br>

###### `ds1`
```{r load_ds1_c, eval = F}
geo_ds1 <- "GSE80157"

g_url_ds1 <- GEOquery::getGEOSuppFiles(geo_ds1, makeDirectory = FALSE, fetch_files = FALSE)
g_file_ds1 <- "../inst/extdata/ds1.txt.gz"
utils::download.file(url = as.character(g_url_ds1$url[2]), destfile = g_file_ds1)

X_ds1 <- read.table(gzfile(g_file_ds1), h=T, sep = '\t', stringsAsFactors = F, row.names = 1)

# convert to rpkm & wb_id
X_ds1 <- RAPToR::format_ids(X_ds1, wormRef::Cel_genes, from = "wb_id", to = "wb_id")
X_ds1 <- raw2rpkm(X = X_ds1, gene.length = wormRef::Cel_genes, id.col = "wb_id", l.col = "transcript_length")





# pheno data
P_ds1 <- Biobase::pData(GEOquery::getGEO(geo_ds1, getGPL = F)[[1]])
P_ds1[,10:34] <- NULL
P_ds1[, 3:8] <- NULL

colnames(P_ds1)[4] <- "strain"
P_ds1$strain <- factor(P_ds1$strain)
P_ds1$title <- gsub('[-\\(\\);]', '.', P_ds1$title)

colnames(X_ds1) <- gsub('RNASeq_riboM_', '', colnames(X_ds1), fixed = T)
P_ds1$title <- gsub('RNASeq_riboM_', '', P_ds1$title, fixed = T)

# get age 
P_ds1$age <- as.numeric(sub('(\\d+)\\shours', '\\1', P_ds1$`time in development:ch1`))


X_ds1 <- X_ds1[, P_ds1$title]

ds1 <- list(g = X_ds1, p = P_ds1)
save(ds1, file = "../inst/extdata/ds1.RData", compress = "xz")

# cleanup
file.remove(g_file_ds1)
rm(geo_ds1, g_url_ds1, g_file_ds1, X_ds1, P_ds1)
```

<br>

###### `ds2`
```{r load_ds2_c, eval = F}
geo_ds2 <- "GSE52861"

g_url_ds2 <- GEOquery::getGEOSuppFiles(geo_ds2, makeDirectory = FALSE, fetch_files = FALSE)
g_file_ds2 <- "../inst/extdata/ds2.txt.gz"
utils::download.file(url = as.character(g_url_ds2$url[2]), destfile = g_file_ds2)

X_ds2 <- read.table(gzfile(g_file_ds2), h=T, sep = '\t', stringsAsFactors = F, row.names = 1)

# convert to rpkm & wb_id
X_ds2 <- RAPToR::format_ids(X_ds2, wormRef::Cel_genes, from = "wb_id", to = "wb_id")
X_ds2 <- raw2rpkm(X = X_ds2, gene.length = wormRef::Cel_genes, id.col = "wb_id", l.col = "transcript_length")


# pheno data
P_ds2 <- Biobase::pData(GEOquery::getGEO(geo_ds2, getGPL = F)[[1]])

# filter relevant fields/samples
P_ds2 <- P_ds2[(P_ds2$`strain:ch1` == 'N2') & (P_ds2$`growth protocol:ch1` == 'Continuous'), ]
P_ds2 <- P_ds2[, c("title", "geo_accession", "time in development:ch1")]

# get age 
P_ds2$age <- as.numeric(sub('(\\d+)\\shours', '\\1', P_ds2$`time in development:ch1`))


# formatting
P_ds2$title <- gsub('RNASeq_polyA_', '', 
                  gsub('hr', 'h', 
                       gsub('-', '.', fixed = T, as.character(P_ds2$title))))
colnames(X_ds2) <- gsub('RNASeq_polyA_','', colnames(X_ds2))
X_ds2 <- X_ds2[, P_ds2$title]

ds2 <- list(g = X_ds2, p = P_ds2)
save(ds2, file = "../inst/extdata/ds2.RData", compress = "xz")

# cleanup
file.remove(g_file_ds2)
rm(geo_ds2, g_url_ds2, g_file_ds2, X_ds2, P_ds2)
```



#### Normalization & Quick look
```{r ex1_load, include=FALSE}
load("../inst/extdata/ds1.RData")
load("../inst/extdata/ds2.RData")
```

```{r ex1_qnorm_log}
ds1$g <- limma::normalizeBetweenArrays(ds1$g, method = "quantile")
ds1$g <- log(ds1$g + 1)

ds2$g <- limma::normalizeBetweenArrays(ds2$g, method = "quantile")
ds2$g <- log(ds2$g + 1)
```

```{r ex1_ql_c1, results='markup'}
ds1$g[1:5, 1:5]

head(ds1$p, n = 5)
```

##### Correlation Matrix
```{r ex1_ql_c2, echo = F, fig.height=5, fig.width=5, out.width="60%"}
cor_ds1 <- cor(ds1$g, method = "spearman")
ord <- order(ds1$p$age)
heatmap(cor_ds1[ord, ord], Colv = NA, Rowv = NA, scale = "none", keep.dendro = F, margins = c(1,5),
        RowSideColors = transp(as.numeric(ds1$p$strain[ord])), labRow = "", labCol = "")
par(xpd = T)
mtext(text = unique(ds1$p$age), side = 1, line = 4, at = seq(-.1,.915, l = 11))

# color key
image(x = c(.95,1), y = seq(0.6,1, l = 10), useRaster = T,
      z = matrix(seq(min(cor_ds1), max(cor_ds1), l = 10), ncol = 10),
      col = hcl.colors(12, "YlOrRd", rev = TRUE), add = T)

strain_legend <- c("let-7", "lin-41", "let-7/lin-41", "N2")
text(.975, 1, pos = 3, labels = expression(rho), font = 2)
text(1, y = seq(0.6,1, l = 10), pos = 4, 
     labels = round(seq(min(cor_ds1), max(cor_ds1), l = 10), 2), cex = .6)

xlp <- 1.025
text(xlp, .5, labels = "Strain", font = 2, cex = .8, adj = .5)
text(xlp, seq(.3,.48, l = 4), labels = strain_legend, adj = 1, pos = 1,
     col = 1:4, font = 2, cex = .7)

mtext(at = xlp, line = 4, side = 1, text = "(hours)", cex = .8)

```

##### Plotting components

```{r ex1_ql_c3}
pca_ds1 <- stats::prcomp(ds1$g, rank = 25)
```

```{r ex1_ql_c4, echo = F, fig.height=6, fig.width=12}
par(mfrow = c(2,4))
invisible(sapply(seq_len(8), function(i){
  plot(ds1$p$age, pca_ds1$rotation[,i], lwd = 2, col = ds1$p$strain,
       xlab = "age", ylab = "PC", main = paste0("PC", i))
  sapply(seq_along(levels(ds1$p$strain)), function(l){
    s <- which(ds1$p$strain == levels(ds1$p$strain)[l])
    points(ds1$p$age[s], pca_ds1$rotation[s,i], col = l, 
           type = 'l', lty = 2)
  })
  # points(ndat$age, pred_ds1_comp[, i], col = "royalblue", type = 'l', lwd = 2)
  if(i == 1)
    legend("topleft", bty = 'n', legend = strain_legend,
           pch = c(rep(1, 4)), lty = c(rep(NA, 4)), col = c(1:4), lwd = 3)
}))
```


### Model fitting

#### Component number

```{r ex1_mf_c1}
nc <- sum(summary(pca_ds1)$importance[3,] < .999) + 1
nc
```

#### Model

```{r ex1_mf_c2}
m_ds1 <- ge_im(X = ds1$g, p = ds1$p, formula = "X ~ s(age, bs = 'cr') + strain", nc = nc)
```

```{r ex1_mf_c3, echo = F}
mp_ds1 <- mperf(scale(ds1$g), predict(m_ds1), is.t = T)
as.data.frame(mp_ds1, row.names = "")
```


#### Validation

##### Predict

```{r ex1_vd_c1, results='markup'}
# setup newdat
n.inter <- 100 # nb of new timepoints
newdat <- data.frame(
  age = seq(min(ds1$p$age), max(ds1$p$age), l = n.inter),
  strain = rep("N2", n.inter) # we want to predict as N2 
  )
head(newdat)

# predict 
pred_m_ds1 <- predict(m_ds1, newdata = newdat)
pred_m_ds1_comp <- predict(m_ds1, newdata = newdat, as.c = TRUE)
```

##### Plot component predictions

```{r ex1_vd_c2, echo = F, fig.width=12, fig.height=6}
par(mfrow = c(2,4))
invisible(sapply(seq_len(8), function(i){
  plot(ds1$p$age, pca_ds1$rotation[,i], lwd = 2, col = ds1$p$strain,
       xlab = "age", ylab = "PC", main = paste0("PC", i))
  sapply(seq_along(levels(ds1$p$strain)), function(l){
    s <- which(ds1$p$strain == levels(ds1$p$strain)[l])
    points(ds1$p$age[s], pca_ds1$rotation[s,i], col = l, 
           type = 'l', lty = 2)
  })
  points(newdat$age, pred_m_ds1_comp[, i], col = "royalblue", type = 'l', lwd = 3)
  if(i == 1)
    legend("topleft", bty = 'n', legend = c("let-7", "lin-41", "let-7/lin-41", "N2", "pred"),
           pch = c(rep(1, 4), NA), lty = c(rep(NA, 4), 1), col = c(1:4, "royalblue"), lwd = c(rep(3,4),4))
}))
```

##### Build reference & stage samples

```{r ex1_vd_c3, eval = F}
# make a 'reference object' 
r_ds1 <- list(interpGE = pred_m_ds1, time.series = newdat$age)

ae_ds1 <- ae(ds1$g, r_ds1$interpGE, r_ds1$time.series)
ae_ds2 <- ae(ds2$g, r_ds1$interpGE, r_ds1$time.series)
```

```{r ex1_vd_c3bis, include = F}
# make a 'reference object' 
r_ds1 <- list(interpGE = pred_m_ds1, time.series = newdat$age)

# speed up w/ no bootstrap
ae_ds1 <- ae(ds1$g, r_ds1$interpGE, r_ds1$time.series, bootstrap.n = 1)
ae_ds2 <- ae(ds2$g, r_ds1$interpGE, r_ds1$time.series, bootstrap.n = 1)
```


```{r ex1_vd_c4, echo=F, fig.height=6, fig.width=12}
par(mfrow = c(1,2))
rg <- range(c(ae_ds1$age.estimates[,1], ds1$p$age))
plot(ae_ds1$age.estimates[,1]~ds1$p$age, 
     xlab = "Chronological age", ylab = "Estimated age (ds1)", 
     xlim = rg, ylim = rg,
     main = "Chron. vs Estimated ages for ds1 (on ds1 reference)", lwd = 2, col = factor(ds1$p$strain))
invisible(sapply(levels(factor(ds1$p$strain)), function(l){
  s <- ds1$p$strain == l
  points(ae_ds1$age.estimates[s,1]~ds1$p$age[s], type = 'l', 
         lty = 2, col = which(l==levels(factor(ds1$p$strain))))
}))

abline(a = 0, b = 1, lty = 3, lwd = 2)
legend("bottomright", legend = c("let-7", "lin-41", "let-7/lin-41", "N2", "x = y"), 
       lwd=3, col=c(1:4, 1), bty='n', pch = c(1,1,1,1,NA), lty = c(rep(NA, 4), 3))

rg <- range(c(ae_ds2$age.estimates[,1], ds2$p$age))
plot(ae_ds2$age.estimates[,1]~ds2$p$age, 
     xlab = "Chronological age", ylab = "Estimated age (ds1)", 
     xlim = rg, ylim = rg,
     main = "Chron. vs Estimated ages for ds2 (on ds1 reference)", lwd = 2)
points(ae_ds2$age.estimates[,1] ~ ds2$p$age, type = 'l', lty = 2)
abline(a = 0, b = 1, lty = 3, lwd = 2)

legend("bottomright", legend = "x = y", lwd=3, col=1, lty = 3, bty='n')
```