---
title: "`RAPToR` - Showcase"
output: 
  rmarkdown::html_document :
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{RAPToR-showcase}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
author: Romain Bulteau
date: "`r format(Sys.Date(), '%B %Y')`"
---

This vignette showcases different uses for `RAPToR`.


<br>
<br>

# Improving Differential Expression analysis with age estimates

Including estimated age as a covariate in Differential Expression (DE) analysis can substantially reduce previously unexplained variation between samples. 
This results in better model fits and consequently better power to detect DE genes. 
In this section, we show that even time-series designs benefit from this approach thanks to unsynchronized replicates.


```{r sc_1, child = "sc_1.Rmd"}

```

<br>
<br>

# Staging samples cross-species 
```{r sc_2, child = "sc_2.Rmd"}

```

<br>
<br>

# Capturing tissue-specific development

```{r sc_3, child = "sc_3.Rmd"}

```

<br>
<br>
<a href="#top">Back to top</a>

<hr>

# References